Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-01-29 ms-3/ms.tex
Draft of para on Figure 2

\begin{Abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see derivation in S1 Text). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.

@@

\section*{Methods}
By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}. To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
\begin{equation}
  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
\end{equation}
where $N$ represents population size, $h_j(t)$ are basis functions, $J$ is the degrees of freedom for the splines, and $\beta_j^{(t)}$ are fitted parameters for a symmetrical window of half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$. 
The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
Explanation of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.

@@

\section*{Results and Discussion}
The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and shrinking toward zero as the effect size increases (Fig.~\ref{fig1}). 
The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$.

@@

Paragraph on results shown in figure 2.

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig2}
\caption{Dispersion analysis of weekly COVID-19 case data for one county. 
a: Weekly incidence in Jefferson County, Alabama. 
b: Estimated dispersion parameter over time. 
c: Comparison of estimated dispersion (grey) with predicted values under the standard model of $\theta_{t+1} = C_{t} / \rho_t$ where $C_t$ represents reported cases and $\rho_t$ the reporting rate, for constant values of $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue). 
This wide range of constant values is intended to enclose the $theta$ values that would be predicted by the standard model under under variable $\rho$.
d: Likelihood ratio test (LRT) statistic over time, highlighting statistically large changes in dispersion (red), defined as p-values less than the Bonferroni-corrected 5\% quantile of a chi-square random variable with one degree of freedom.}\label{fig2}
\end{figure}
\clearpage

>>>>

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.

@@

\section*{Methods}
By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}. To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
\begin{equation}
  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
\end{equation}
where $N$ represents population size, $h_j(t)$ are basis functions, $J$ is the degrees of freedom for the splines, and $\beta_j^{(t)}$ are fitted parameters for a symmetrical window of half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$. 
The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
Explaination of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.

@@

\section*{Results and Discussion}
The simulation studies indicate our LRT framework accurately detects changes in dispersion, with p-values averaging 0.5 when the effect size is zero and decreasibg toward zero as the effect size increases (Fig.~\ref{fig1}). 
The framework is also robust to the range of population sizes we addressed in the empirical data---county population sizes ranged from approximately 48 thousand to 9.9 million, and we tested the framework on simulated populations between 10 thousand and 10 million, yielding accurate estimates within the range of $\theta$ values, $10^-2 \leq \theta \leq 10^2$, which encompasses all operationally relevant values of $\theta$ for COVID-19 incidence data and many other infectious diseases: values lower than this range would focus the PMF for cases almost entirely on 0 while values higher than this range are effectively Poisson.

@@

Applying the method to COVID-19 cases in US counties enabled investigation of changes in the $\theta$ in relation to both observed trends and model expectations (Figure 2). 
Periods of increased case count variability corresponded with decreases in the dispersion parameter, indicating that $\log_{10}\theta$ was dynamic and displayed both expected and unexpected patterns of variation. 
In some instances, $\theta$ appeared to vary inversely with incidence, aligning with standard epidemic theory, while in other periods, deviations from that expectation occurred, potentially highlighting shifts in the underlying transmission dynamics. 
Notably, significant changes in the dispersion parameter were observed during major epidemic transitions, such as the surge at the beginning of 2022, coinciding with heightened volatility in reported cases, and at the end of the time series, during the approach to endemicity as the landscape of susceptibility was changing and potentially becoming more complex~\citep{lavine2021immunological}.
In other words, as a larger portion of incidence involved repeat infections (albiet potentially with different variants), prior immunity became a larger factor shaping epidemic dynamics, and potentially impacting the demography of burden, in particular the burden as it depends on age, and potentially particularly in children~\citep{lavine2021immunological}.
These findings underscore the complex behavior of the dispersion parameter, which both varied with changes in case count regimes and highlighted departures from the model expectations described by Eq. 1.

\clearpage
\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig2}
\caption{Dispersion analysis of weekly COVID-19 case data for one county. 
Analysis for all counties together is shown in Fig 3.
a: Weekly incidence in Jefferson County, Alabama. 
b: Estimated dispersion parameter over time. 
c: Comparison of estimated dispersion (grey) with predicted values under the standard model of $\theta_{t+1} = C_{t} / \rho_t$ where $C_t$ represents reported cases and $\rho_t$ the reporting rate, for constant values of $\rho_t = 0.1$ (black) and $\rho_t = 0.9$ (blue). 
This wide range of constant values is intended to enclose the $theta$ values that would be predicted by the standard model under under variable $\rho$.
d: Likelihood ratio test (LRT) statistic over time, highlighting statistically large changes in dispersion (red), defined as p-values less than the Bonferroni-corrected 5\% quantile of a chi-square random variable with one degree of freedom.}\label{fig2}
\end{figure}
\clearpage
